---
title: Timeout on parallel threads in R
description: Python's `multiprocessing` and `threading` libraries both have a timeout parameter for re-joining threads after they've finished. This provides an easy way to launch multi-threaded jobs while ensuing that no single thread exceeds a specified timeout. This post describes one way to implement an equivalent functionality in R.
date: 12/01/2022
link: blog/blog011.html
---

# Timeout on parallel threads in R

Python's
[`multiprocessing`](https://docs.python.org/3/library/multiprocessing.html) and
[`threading`](https://docs.python.org/3/library/threading.html) libraries both
have a timeout parameter for re-joining threads after they've finished. This
provides an easy way to launch multi-threaded jobs while ensuing that no single
thread run for longer than a specified timeout. This is very useful in
implementing a standard "timeout on a function call" operation, as detailed in
[this Stack Overflow question of that
title](https://stackoverflow.com/questions/492519/timeout-on-a-function-call)
which offers a bewildering variety of approaches to that problem. Among the
easiest of those is [the recommendation to rely on the `multiprocessing`
libraries's `join()` operation](https://stackoverflow.com/a/14924210) which
accepts a `timeout` parameter, [as described in the library's
documentation](https://docs.python.org/3/library/multiprocessing.html#multiprocessing.Process.join).
There is also an equivalent parameter [for python's other main parallelisation
library,
`threading`](https://docs.python.org/3/library/threading.html#threading.Thread.join).

A nice example of the usefulness of this `timeout` parameter in action is given
in [the `fitter` package](https://github.com/cokelaer/fitter) by
[&#64;cokelaer](https://github.com/cokelaer) for fitting probability
distributions to observed data. The main function fits a wide range of
different distributions, and can even automagically select the best
distribution according to specified criteria. This is done through fitting
different distributions in parallel on different threads, generally greatly
speeding up calculations. Distributional fitting is, however, often an
iterative procedure, meaning the duration required to generate a fit within
some specified tolerance can not be known in advance. Parallel threads by
default must wait for all to terminate before individual results can be joined.
To ensure distributional fits are generated within a reasonable duration,
[`fitter` has a `_timed_run`
function](https://github.com/cokelaer/fitter/blob/cf222aab741492917bd3a2d1af821e0b5344907d/src/fitter/fitter.py#L429)
to:

> spawn a thread and run the given function ... and return the given default
> value if the timeout is exceeded.

The bit of that function which controls the timeout consists of the following
lines (with code for exception handling removed here):

```{python, eval = FALSE}
def _timed_run (self, func, args=()):
    class InterruptableThread(threading.Thread):
        def __init__(self):
            threading.Thread.__init__(self)
            self.result = default

        def run(self):
            self.result = func(args)

    it = InterruptableThread()
    it.start()
    it.join(self.timeout)
    return it.result
```

That represents a succinct way to run a multi-threaded job in which each thread
obeys a specified timeout parameter. There is no equivalent of this
functionality within any of the parallelisation packages in R, and so this post
describes one approach to implementing equivalent functionality.

## Timeout on parallel jobs in R

R offers a wide variety of packages for parallel processing, from the
[`parallel`
package](https://stat.ethz.ch/R-manual/R-devel/library/parallel/doc/parallel.pdf)
that is part of R's "recommended" software included with every distribution
(the equivalent of Python's standard library), to the one-stop shop of [the
`future` package](https://future.futureverse.org/). Almost all contributed
libraries like `future` nevertheless have two common attributes relevant in
this context:

1. They rely on the underlying functionality of the `parallel` package; and
2. They offer no equivalent of Python's `timeout` parameter.

There is nevertheless one R package which does offer precisely this timeout
functionality: [the `callr` package by Gábor Csárdi and
Winston Chang](https://callr.r-lib.org) for "calling R from R" - that is, for,

> performing computation in a separate R process, without affecting the current R process

### callr's timeout parameter

The `callr` package offers two main modes of calling processes: [as blocking,
foreground processes via
`callr::r()`](https://callr.r-lib.org/reference/r.html), or [as non-blocking,
background processes via
`callr::r_bg()`](https://callr.r-lib.org/reference/r_bg.html). The foreground
`r()` function has an explicit `timeout` parameter, which returns a
`system_command_timeout_error` if the specified timeout (in seconds) is
exceeded. The following code demonstrates this functionality, wrapping the main
call in `tryCatch()` to process the timeout errors:

```{r long-fn-fg}
timeout_fn <- function (x = 1L, timeout = 10) {
    tryCatch (
        callr::r (fn, args = list (x = x), timeout = timeout),
        error = function (e) NA
        )
}
```

The following code then constructs an arbitrarily slow function and passes it
to our timeout function:

```{r slow-fn1, eval = FALSE}
fn <- function (x = 10L) {
    vapply (seq (x), function (i) {
                Sys.sleep (0.2)
                runif (1)
        }, numeric (1))
}
system.time (
    x <- timeout_fn (x = 10, timeout = 2)
    )
x
```
<br>
```{r slow-fn1-out, echo = FALSE}
fn <- function (x = 10L) {
    vapply (seq (x), function (i) {
                Sys.sleep (0.2)
                runif (1)
        }, numeric (1))
}
system.time (
    x <- timeout_fn (x = 10, timeout = 2)
    )
x
```
<br>

That function timed out as expected, because the requested `x = 10` values,
each taking 0.2s, equals the specified timeout value. (The overhead associated
with all the other code lines is enough to push the total calculation beyond
that limit.) Compare what happens when the `timeout` is extended well beyond
that limit:

```{r slow-fn2, eval = FALSE}
system.time (
    x <- timeout_fn (x = 10, timeout = 10)
    )
x
```
<br>
```{r slow-fn2-out, echo = FALSE}
system.time (
    x <- timeout_fn (x = 10, timeout = 10)
    )
x
```

The `timeout` parameter of `callr::r()` can thus be used to directly implement
a timeout parameter. The following sub-section demonstrates how to extend this
to parallel jobs.

### Timeout in parallel jobs.

The following code uses the `mclapply` function of the `parallel` package to
call the `timeout_fn` a specified number of times, with different values of `x`
each time.

```{r parallel, eval = FALSE}
set.seed (1)
n <- sample (1:20, size = 10, replace = TRUE)
nc <- parallel::detectCores () - 1L
system.time (
    res <- parallel::mclapply (mc.cores = nc, n, function (i)
                               timeout_fn (x = i, timeout = 2))
    )
```
<br>

```{r sys.time, echo = FALSE}
c (user = 1.754, system = 0.544, elapsed = 3.008)
```
<br>

```{r parallel-out, eval = FALSE}
print (res)
```
<br>

```{r parallel-out-for-real, echo = FALSE}
res <- as.list (rep (NA, 10L))
res [[1]] <- c (0.20134728, 0.09508085, 0.75240848, 0.30041337)
res [[2]] <- c (0.5837042, 0.6133771, 0.3121486, 0.2943205, 0.4455983, 0.5102744, 0.8867751)
res [[3]] <- 0.9381157
res [[4]] <- c (0.9201705, 0.9656466)
res [[9]] <- 0.7515151
print (res)
```

And that took 3 seconds to calculate 10 jobs, of which 5 timed out at the
specified time of 2 seconds. That demonstrates how the functionality of the
`callr` package can be combined with R's parallelisation routines to achieve
the functionality, if not the elegance, of Python's `multiprocessing` and
`threading` libraries.

### Timeout parameters and 'future' packages

Processes triggered by the `callr` package do not generally play nicely with
the core `future` package, which was likely one motivation for Henrik Bengtsson
to develop [the `future.callr` package](https://future.callr.futureverse.org/)
which explicitly uses `callr` to run each process. The processes are
nevertheless triggered as `callr::r_bg()` processes which do not have a
`timeout` parameter. While it is possible to directly implement a timeout
parameter of `r_bg` processes by monitoring until timeout and then using the
`kill` method, the `future.callr` package does not directly expose the `r_bg`
processes necessary to enable this. There is therefore currently no safe way to
implement a timeout parameter along the lines demonstrated here within any
`futureverse` packages.

