---
title: The troubles with getting help files in R
description: A primer on ways to extract the actual content of help files. Because one day people will hopefully start text-mining these things, and show us all sorts of things we never knew about the people who make R packages. When they do, this entry will hopefully help.
date: 29/09/2020
link: blog/blog008.html
---

# The troubles with getting help files in R

## Databases of help files in R

R has a very well structured system for documenting and accessing help for
packages. In most systems, attempts to access help files will result in
a dedicated window opening up with nicely formatted help content for
a requested topic. This blog entry addresses the issue of how to extract the
underlying text of those files, for example in order to do any kind of text
mining-type analyses.

The content of the help files can be extracted for a given package via the
[`tools::Rd_db()`
function](https://stat.ethz.ch/R-manual/R-devel/library/tools/html/Rdutils.html).
That function works like this:

```{r rd_db}
x <- tools::Rd_db (package = "tools")
class (x)
length (x)
class (x [[1]])
```

Say I want to extract the help file shown on the `html` page for `Rd_db` linked
to immediately above. Then I just have to find the entry in the `Rd_db` data.
As a first try, I simply examine the names of the files in the database, and
try to match (via `grep`) the one called something like `rd_db`:

```{r rd_db2}
grep ("rd_db", names (x), ignore.case = TRUE)
```

The database contains no file called `Rd_db` or the like. If you click again on
the above link to the html entry you'll notice that the page itself is called,
`Rdutils`. Where does that name come from? Help files for R packages are
contained within a `/man` directory of the package source. When a package is
installed, all files within that directory (which end with a suffix of `.Rd`)
are compiled into a binary database object which can then be read by the
[`Rd_db()`
function](https://stat.ethz.ch/R-manual/R-devel/library/tools/html/Rdutils.html).
So the databases of help files for any given package contain one entry for each
file in the original `/man` directory of the package source, with the names of
those original files transferred over to the names of the corresponding entries
in the `Rd_db` file. These databases in installed packages are no longer
contained within directories named `/man`, rather they are compiled within
a directory called `/help`. The contents of this directory can be readily
examined with code like the following:


```{r}
loc <- file.path (R.home (), "library", "tools", "help")
list.files (loc)
```

And the two `tools.rdb` and `toools.rdx` files represent the binary database of
help files for the `tools` package. An alternative way to access the databases
contained within that directory is via the [`lazyLoad()`
function](https://stat.ethz.ch/R-manual/R-devel/library/base/html/lazyload.html).
(And clicking on that website indicates another inconsistency that the function
is called `lazyLoad`, yet the page is named `lazyload`, for reasons which
should become clear as you read on.)

```{r}
package <- "tools"
loc <- file.path (R.home (), "library", package, "help", package)
e <- new.env ()
chk <- lazyLoad (loc, envir = e)
head (names (x))
head (ls (envir = e))
```

Those last two commands reveal that the entries in the object returned from
`Rd_db()` are the original and full file names within the `/man` directory of
the package source, while the corresponding names when `lazyLoad`ed have the
suffix, `.Rd`, removed. The following line nevertheless confirms that the two
methods yield identical results:

```{r rd-vs-lazyload}
all (ls (envir = e) == gsub ("\\.Rd$", "", names (tools::Rd_db (package = "tools"))))
```

### An alternative approach

An alternative approach to extract some of the information contained in the
`Rd_db` object uses a trick that the [`utils::help`
function](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/help.html)
can be called without specifying a `topic`. Additionally specifying `help_type
= "text"` will then retrieve a few components of the database in text form.

```{r}
package <- "tools"
h <- utils::help (package = eval (substitute (package)), help_type = "text")
class (h)
```
At that point, attempting to `print` the object `h` will simply open the help
file the usual way, rather than giving you the textual content. Noting the
output of the following,
```{r}
str (h)
```
leads to the obvious next step of examining
```{r hinfo-fakey, eval = FALSE, echo = TRUE}
h$info
```
```{r hinfo, eval = TRUE, echo = FALSE}
hcut <- lapply (h$info, function (i) i [1:10])
hcut
```

And the second component of the `h$info` object has the names and descriptions
of each entry in the help database.

## Getting help content for a particular function

If we want to analyse the textual content of help files, then we obviously need
a way to extract that content for any given function. Armed with the basics
described above, let's say we want to extract the content of the help file for
the [`tools::Rd_db()`
function](https://stat.ethz.ch/R-manual/R-devel/library/tools/html/Rdutils.html).
If you click on that link, you'll notice that the page which describes only the
function `Rd_db()` is actually called, `Rdutils`. So how could we automatically
extract the content of the help file for `Rd_db()`, or indeed any particular
function, when the help files describing our desired function may have entirely
arbitrary names?

The full entry for `Rdutils` looks like this:

```{r}
x [["Rdutils.Rd"]]
```

And you'll notice at the top that `Rd_db` is given as an `alias`. The structure
of these files is described in a section of the ["Writing R Extensions"
manual](https://cran.r-project.org/doc/manuals/R-exts.html#Documenting-functions),
which explains that these files contain a "name", a "title", and optional
"alias" entries. Comparing the above text to the formatted `html` help page for
[`Rd_db()`](https://stat.ethz.ch/R-manual/R-devel/library/tools/html/Rdutils.html)
reveals what those three fields are:

1. The "name" field defines the name of a single help topic, which may or may not
   be the name of the original `/man` directory file in the package source
   (more on this below).
2. The "title" field specifies an arbitrary description which will appear at
   the top of the help file.
3. The "alias" fields specify topics which will be linked to the given help
   file.

So to locate the help entry for a nominated function, we need to find match that
function with an `alias` entry for some help file which we do not necessarily
know the name of. As long as we know the package in which we are searching, we
can then simply extract all `alias` entries for every single help file. The
example in the help file for the
[`Rd_db()`](https://stat.ethz.ch/R-manual/R-devel/library/tools/html/Rdutils.html)
function use a non-exported function called `.Rd_get_metadata()` (non-exported
meaning that function can only be called via the triple-colon method as
`tools:::.Rd_get_metadata()`, and also meaning that there will be no help entry
for this function). This function can be used to extract all "alias" fields for
every help topic:

```{r get-aliases}
aliases <- lapply (x, function (i) tools:::.Rd_get_metadata (i, "alias"))
```

Code like the following can then be used to find the file which describes the
`Rd_db` function.

```{r find-aliases}
myfn <- "Rd_db"
aliases [which (vapply (aliases, function (i) myfn %in% i, logical (1)))]
```

And that gives us the name of the help file describing our desired function.

## Getting help content for a particular function (#2)

An alternative approach to finding the names of help files associated with
a specified function is to use the [`help.search()`
function](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/help.search.html)
which returns the following kinds of results:

```{r hs-fakey, echo = TRUE, eval = FALSE}
hs <- help.search (pattern = "Rd_db", package = "tools")
str (hs)
hs$matches
```
```{r hs, echo = FALSE, eval = TRUE}
hs <- help.search (pattern = "Rd_db", package = "tools")
hs$lib.loc [1] <- "/<path>/<to>/<my>/<lib>/<loc>"
str (hs)
```
```{r hs-matches}
hs$matches
```

We can see there that the final, `"Entry"` column includes `Rd_db`, and
specifies that it is an `"alias"`. The name of the associated file is also
given there as "Rdutils".


## Names of help topics; names of help files

I indicated above that the "name" field of an "Rd" file 

> defines the name of a single help topic, which may or may not be the name of
the original `/man` directory file in the package source

An example of a help topic which differs from the name of the underlying file
arises courtesy of the `"formatC"` function from the "base" package.

```{r}
hs <- help.search (pattern = "formatC", package = "base")
hs$matches [hs$matches$Topic == "formatC", ]
```

So "formatC" is the official name of one of the help topics within the "base"
package, and therefore should also be the name of the entry within its help
database. And yet look what happens when the help database is accessed via
`lazyLoad`:

```{r}
package <- "base"
loc <- file.path (R.home (), "library", package, "help", package)
e <- new.env ()
chk <- lazyLoad (loc, envir = e)
fns <- ls (envir = e)
fns [grep ("formatC", fns, ignore.case = TRUE)]
```

And the entry in the database is called `formatc` (lower-case "c"), yet the
content of that entry declares a "name" of `formatC` (upper-case "C"). So the
"Name" entry in the object returned by the [`help.search()`
function](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/help.search.html)
is not actually the name
of the `/man` file in the source package, rather it is the "name" entry
specified in the actual "Rd" file.

The contents of the corresponding file can nevertheless be extracted via
`Rd_db` like this:

```{r}
x <- tools::Rd_db (package = "base")
aliases <- lapply (x, function (i) tools:::.Rd_get_metadata (i, "alias"))
myfn <- "formatC"
i <- which (vapply (aliases, function (i) myfn %in% i, logical (1))) # number of entry in db
names (x) [i]
tools:::.Rd_get_metadata (x [[i]], "name")
```

This indicates that the `help.search()` function ought not be used to extract
the contents of help files, but that the `Rd_db()` function can be used as
illustrated above, with no need to worry about the names of the underlying
files. A final function might look something like the following:

```{r}
help_text <- function (fn_name, package) {
    x <- tools::Rd_db (package = package)
    aliases <- lapply (x, function (i) tools:::.Rd_get_metadata (i, "alias"))
    i <- which (vapply (aliases, function (i) fn_name %in% i, logical (1)))
    return (x [[i]])
}
```



### Conclusion

Hopefully this has been helpful to anyone wanting to extract the actual
contents of R help files. While the objects extracted by the methods described
above can generally be treated as `character` objects, and any form of parsing
applied, they are also objects with a defined class of `Rd`. There are several
methods already available to parse such objects, in particular those described
in the help file for the [`parse_Rd()`
function](https://stat.ethz.ch/R-manual/R-patched/library/tools/html/parse_Rd.html)
which claims at the outset that,

> This function parses ‘Rd’ files according to the specification given in [https://developer.r-project.org/parseRd.pdf](https://developer.r-project.org/parseRd.pdf)

The document referred to there goes in to extensive detail about methods for
parsing these objects.

