---
title: What are matrices in R?
description: If everything in R is a vector, then what is a matrix? This entry
    will demonstrate that even matrices are vectors, and that processing of
    matrices can in certain circumstances be considerably more efficient if
    they are treated as simple vectors.
date: 31 July 2019
link: blog/blog005.html
---

# What are matrices in R?

"R is a shockingly dreadful language for an exceptionally useful data analysis
environment" (
<a target="_blank" rel="noopener noreferrer" href="http://arrgh.tim-smith.us/">
Tim Smith & Kevin Ushey
</a>). One of the strangest manifestations of claims like these is that, 
<a target="_blank" rel="noopener noreferrer" href="https://www.noamross.net/blog/2014/4/16/vectorization-in-r--why.html">
"Everything in R is a vector"
</a>. The simple question that then arises is, What is a matrix? One commonly cited current repository of things R is Hadley Wickham's book, 
<a target="_blank" rel="noopener noreferrer" href="http://adv-r.had.co.nz">
Advanced R
</a>, which has a section on 
<a target="_blank" rel="noopener noreferrer" href="http://adv-r.had.co.nz/Data-structures.html">
Data Structures
</a> which simply states that a matrix is the two dimensional equivalent of a vector, and that,
<a target="_blank" rel="noopener noreferrer" href="http://adv-r.had.co.nz/Data-structures.html#matrices-and-arrays">
"Adding a `dim` attribute to an atomic vector allows it to behave like a multi-dimensional array."
</a> The chapter linked to above goes on to say that, "Vectors are not the only
1-dimensional data structure. You can have matrices with a single row or single
column, or arrays with a single dimension. They may print similarly, but will
behave differently. The differences aren’t too important." This blog entry will
attempt to illustrate the kind of circumstances under which differences between
vectors and matrices actually become quite important indeed.

## An initial illustration

Vectors do differ from matrices, as the following code clearly illustrates:
```{r, echo = FALSE}
library (magrittr)
```

```{r rowsum1-fakey, eval = FALSE}
n <- 1e6
x <- runif (n)
y <- runif (n)
xy <- cbind (x, y) # a matrix
rbenchmark::benchmark (
                       res <- x + y,
                       res <- rowSums (xy),
                       replications = 100,
                       order = NULL) [, 1:4]
```
```{r rowsum1, echo = FALSE}
n <- 1e6
x <- runif (n)
y <- runif (n)
xy <- cbind (x, y) # a matrix
knitr::kable (rbenchmark::benchmark (
                                     res <- x + y,
                                     res <- rowSums (xy),
                                     replications = 100,
                                     order = NULL) [, 1:4])
```

Adding the two rows of a matrix takes 3-4 times longer than adding two
otherwise equivalent vectors. And okay, that's very likely something to do with
the `rowSums` function rather than the matrix itself, but why should these two
behave so differently? At that point, I must freely admit to being not
sufficiently clever to have uncovered the actual reason in the 
<a target="_blank" rel="noopener noreferrer" href="https://github.com/wch/r-source/blob/trunk/src/main/array.c">
underlying C source code.
</a> The answer must lie somewhere in there, so any pointers would be greatly
appreciated. Short of that, the following is a phenomenological explanation,
derived through attempting to reconstruct in C code what `rowSums` is actually
doing.

Direct vector addition must work something like the following C code, written
here in a form able to be directly parsed in R via the 
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/package=inline">
inline package
</a>.

```{r add1}
library (inline)
add <- cfunction(c(a = "numeric", b = "numeric"), "
                 int n = length (a);
                 SEXP result = PROTECT (allocVector (REALSXP, n));
                 double *ra, *rb, *rout;
                 ra = REAL (a);
                 rb = REAL (b);
                 rout = REAL (result);
                 for (int i = 0; i < n; i++)
                     rout [i] = ra [i] + rb [i];
                 UNPROTECT (1);

                 return result;
                 ")
```
That's a simple C function to add two vectors and return the result, with most
of the code providing the necessary scaffolding for an R function. The
following benchmark compares that with the previous two equivalent functions.
```{r benchmark1-fakey, eval = FALSE}
rbenchmark::benchmark (
                       res <- x + y,
                       res <- add (x, y),
                       res <- rowSums (xy),
                       replications = 100,
                       order = NULL) [, 1:4]
```
```{r benchmark1, echo = FALSE}
knitr::kable (rbenchmark::benchmark (
                                     res <- x + y,
                                     res <- add (x, y),
                                     res <- rowSums (xy),
                                     replications = 100,
                                     order = NULL) [, 1:4])
```

So our `add` function is broadly equivalent to R's underlying code for vector
addition, and correspondingly, considerably more efficient than `rowSums`
applied to an equivalent matrix. This naturally fosters the question of whether
the inefficiency arises in `rowSums` itself, or whether it is somehow something
inherent to R's internal representation of matrices and/or matrix operations?
The following code provides an initial answer to that quesiton.
```{r benchmark2-fakey, eval = FALSE}
rbenchmark::benchmark (
                       res <- x + y,
                       res <- add (x, y),
                       res <- rowSums (xy),
                       res <- xy [, 1] + xy [, 2],
                       replications = 100,
                       order = NULL) [, 1:4]
```
```{r benchmark2, echo = FALSE}
knitr::kable (rbenchmark::benchmark (
                                     res <- x + y,
                                     res <- add (x, y),
                                     res <- rowSums (xy),
                                     res <- xy [, 1] + xy [, 2],
                                     replications = 100,
                                     order = NULL) [, 1:4])
```

And direct addition of two columns of a matrix, through indexing into those
columns, is roughly as *inefficient* as `rowSums` itself, while direct addition
of the equivalent vectors remains 3-4 times more efficient.

### How are matrices stored?

So the reason for the relative inefficiency of `rowSums` is likely to extend
directly from the column selection operation, `xy[, i]`. The reference manual
for the C-level details of data storage and sub-selection in R is the online
compendium,
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/doc/manuals/r-release/R-ints.html#SEXPs">
R Internals
</a>, yet even this has remarkably little to say in regard to how matrices are
actually stored or manipulated. The key is a single incidental statement that, 
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/doc/manuals/R-ints.html#Large-matrices">
"Matrices are stored as vectors"
</a>. The storage can then be understood through reading the details of vector
storage, and then simply figuring out how the indexing of a matrix-as-vector is
implemented. This can be easily discerned from direct conversion within R:
```{r matrix-as-vector}
as.vector (cbind (1:5, 6:10))
```
The columns of a matrix are directly concatenated within the vector object.
This enables us to then re-write the above C code for vector addition to
instead accept a matrix object, noting that the indices `i` and `n + i`
respectively refer to the first and second columns of the matrix.
```{r}
matadd <- cfunction(c(a = "numeric"), "
                 int n = floor (length (a) / 2);
                 SEXP result = PROTECT (allocVector (REALSXP, n));
                 double *ra, *rout;
                 ra = REAL (a);
                 rout = REAL (result);
                 for (int i = 0; i < n; i++)
                     rout [i] = ra [i] + ra [n + i];
                 UNPROTECT (1);

                 return result;
                 ")
```

Benchmarking that against the previous versions, and including an additional
comparison of direct matrix addition, gives the following results.
```{r benchmark3-fakey, eval = FALSE}
rbenchmark::benchmark (
                       res <- x + y,
                       res <- add (x, y),
                       res <- rowSums (xy),
                       res <- xy [, 1] + xy [, 2],
                       res <- matadd (xy),
                       res <- xy + xy,
                       replications = 100,
                       order = NULL) [, 1:4]
```
```{r benchmark3, echo = FALSE}
knitr::kable (rbenchmark::benchmark (
                                     res <- x + y,
                                     res <- add (x, y),
                                     res <- rowSums (xy),
                                     res <- xy [, 1] + xy [, 2],
                                     res <- matadd (xy),
                                     res <- xy + xy,
                                     replications = 100,
                                     order = NULL) [, 1:4])
```

That benchmark demonstrates that operations on matrix columns are only as
efficient as equivalent operations on vectors when the matrices are treated as
singular vector objects. Direct addition of entire matrices (`xy
+ xy`) is also as efficient as vector addition, taking here around twice as
long because twice as many values are being added. Inefficiencies arise in
handling matrices only when extracting individual rows or columns -- the `xy[,
i]` operations, presumably because these operations involve creating an
additional copy of the entire row or column. 

## Conclusion

What the above code was intended to demonstrate was that matrices should only
be considered to be **like** vectors in the sense of operations on the entire
objects. Sub-setting or sub-selecting of matrices involves creating additional
copies of the sub-set/sub-selected portions, and is comparably less efficient
than equivalent vector operations. In particular, efficient C or C++ operations
on matrices should index directly into the underlying vector object, rather than
sub-setting particular rows or columns of the matrices.

The assertion that everything in R is a vector hereby deepens: Even matrices in
R are vectors, and should in many circumstances be treated as such.

