---
title: The troubles with getting help files in R
description: How to extract the actual content of help files.
date: 08/09/2020
link: blog/blog008.html
---

# The troubles with getting help files in R

R has a very well structured system for documenting and accessing help for
packages. This entry describes how to extract the actual content of help files.
The `tools` package has a few functions for working with help files, starting
with
[`parse_Rd`](https://stat.ethz.ch/R-manual/R-patched/library/tools/html/parse_Rd.html)
which claims at the outset that,

> This function parses ‘Rd’ files according to the specification given in [https://developer.r-project.org/parseRd.pdf](https://developer.r-project.org/parseRd.pdf)

The document referred to there tells you everything you maybe did not ever
really want to know about *how* the `parse_Rd()` function actually does the
parsing, yet not anything about how the mysterious `.Rd` objects supposedly
parsed by the function are actually accessed in the first place. Absent that
information, it remains very difficult to figure out how to use `parse_Rd`.

## Extracting a help database in text form

Help files for R packages (since around version 2 at least) are stored in
a standard location, within the `"help"` directory of the local package
installation. The following code demonstrates for the `stats` package, which
includes the `rnorm()` function. Note that when you type `?rnorm` (or
`help(rnorm)`), the help page that opens up includes details for `rnorm`,
`pnorm`, `dnorm`, and `qnorm`. Also note that the help page itself is named
`"Normal"` (generally in the top left corner). This is what is referred to as
the
["topic"](https://cran.r-project.org/doc/manuals/R-exts.html#Documenting-functions),
and is not necessarily the name of the function for which help is sought.

```{r tools-pkg}
loc <- file.path (R.home (), "library", "stats", "help")
list.files (loc)
```

That directory does not contain any actual `.Rd` objects, or anything that can
be submitted directly to `parse_Rd()`. And yet it does contain all of the help
information, in the format of one `.rdb` and one `.rdx` file. As explained in
one answer to [this StackOverflow question](
https://stackoverflow.com/questions/46184224/how-to-open-rdb-file-using-r),
these files can be loaded with the [`lazyload()`
function](https://stat.ethz.ch/R-manual/R-devel/library/base/html/lazyload.html).
Armed with that knowledge, the following function illustrates one simple way to
extract the contents of help files as text entries.

```{r}
gethelp <- function (package, fn) {
    loc <- file.path (R.home (), "library", package, "help", package)
    e <- new.env ()
    chk <- lazyLoad (loc, envir = e)
    get (fn, envir = e)
}
x <- gethelp ("stats", "Normal")
class (x)
print (x)
```

So `lazyLoad()` can be used to directly access the `.Rd` objects containing the
help files, but you first need to know the topic. If we had not already noted
that the topic for `rnorm` was in fact `"Normal"`, then we could of course try
`gethelp ("stats", "rnorm")`, but that would just produce an error because
there is no topic called "rnorm". So is there any direct way to find the text
context of the help topic for "rnorm"?


## Digging deeper

An alternative approach uses a trick that the [`utils::help`
function](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/help.html)
can be called without specifying a `topic`. Additionally specifying `help_type
= "text"` will then retrieve the entire database in text form.

```{r}
pacakge <- "stats"
h <- utils::help (package = eval (substitute (pacakge)), help_type = "text")
class (h)
```
At that point, attempting to `print` the object `h` will simply open the help
file the usual way, rather than giving you the textual content. Noting the
output of the following,
```{r}
str (h)
```
leads to the obvious next step of examining
```{r hinfo-fakey, eval = FALSE, echo = TRUE}
h$info
```
```{r hinfo, eval = TRUE, echo = FALSE}
hcut <- lapply (h$info, function (i) i [1:10])
hcut
```
which (shown above in artificially condensed form) is indeed a simple
`character` vector containing the names of all of the help topics contained
within the database. So that gives us a list of all of the topics, but we will
do not know how to match `rnorm` on to its associated topic of `"Normal"`. One
approach for finding topics associated with a specified function is to use the
[`help.search()`
function](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/help.search.html)
which returns the following kinds of results:

```{r hs-fakey, echo = TRUE, eval = FALSE}
hs <- help.search (pattern = "rnorm", package = "stats")
str (hs)
hs$matches
```
```{r hs, echo = FALSE, eval = TRUE}
hs <- help.search (pattern = "rnorm", package = "stats")
hs$lib.loc [1] <- "/<path>/<to>/<my>/<lib>/<loc>"
str (hs)
hs$matches
```

We can see there that the final, `"Entry"` column includes `rnorm`, and
specifies that it is an `"alias"`. The associated `"Topic"` can thus be found
simply as,

```{r}
hs$matches [hs$matches$Entry == "rnorm", ]
```

And the `"Topic"` for the `"alias"` of `rnorm` is "Normal".

## What about extracting all help topics for a package?

That procedure works for extracting the textual contents of the help file for
any specified function. What if we want to extract entries for all functions
for a given package? Packages may have many more functions that help topics, as
with the example considered here where one topic describes four functions. We
can list the functions for a package like this:

```{r}
fns <- ls ("package:stats")
```

However, that list includes not only functions, but methods, operators, and
a variety of other possible things. The variety can be seen with the following
code:
```{r}
fn_classes <- vapply (fns, function (i)
                      class (get (i, envir = as.environment ("package:stats"))) [1],
                      character (1))
table (fn_classes)
```
We can use that to narrow the listed objects from a package down to functions
only. We can then use the above results from `help.search` for each of those
functions to find the associated `"Topic"` entry, and thus to extract the help
file in textual form.

## A final word of caution

A final warning, by way of demonstrating how messy the business of extracting
the textual content of help files can get, comes courtesy of the help topic
`"FDist"`, again from the "stats" package.

```{r}
hs <- help.search (pattern = "FDist", package = "stats")
hs$matches [hs$matches$Topic == "FDist", ]
```

So "FDist" is the official name of one of the help topics within the "stats"
package, and therefore should also be the name of the entry within its help
database. And yet look what happens via the first of the above approaches:

```{r}
loc <- file.path (R.home (), "library", package, "help", package)
e <- new.env ()
chk <- lazyLoad (loc, envir = e)
fns <- ls (envir = e)
fns [grep ("FDist", fns, ignore.case = TRUE)]
```
And the entry in the database is called "Fdist", yet the content of that entry
declares a topic of "FDist". This mismatch means that the previous `get_help()`
function will fail on the latter, and only work on the mis-typed version of
"Fdist". It is also manifest in the official web page for that topic having the
address
[https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Fdist.html](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Fdist.html),
with the same address using "FDist" **not** leading to the desired page, even
though the first line of the resultant page clearly informs you that you're
looking at a paged called "FDist" rather than "Fdist". I could dig further into
where the actual names of the entries within the help databases come from, but
that's likely enough to show how non-trivial this whole thing quickly becomes.
But hopefully also helpful to anyone wanting to delve into the actual contents
of R help files.


