---
title: Caching via background R processes
description: Caching is implemented because it saves time, generally by saving
    the results of one function call for subsequent reuse. Background processes
    are also commonly implemented as time-saving measures, through delegating
    long-running tasks to "somewhere else", allowing you to keep focussing on
    whatever (un)important things you were doing in the meantime. This blog
    entry describes how to combine the two to save double time through caching
    via background processes.
date: 28 May 2019
link: blog/blog003.html
---

# Caching via background R processes

The title of this blog entry should be fairly self-evident for those who might
incline to read it, yet is motivated by the simple fact that there currently
appear to be no online sources that clearly describe the relatively
straightforward process of using background processes in **R** to cache
objects. (Check out search engine results for 
<a target="_blank" rel="noopener noreferrer" href="https://duckduckgo.com/?q=caching+background+R+processes&t=ffab&ia=web">
"caching background R processes"
</a>: most of the top entries are for Android, and even opting for other search
engines 
<a target="_blank" rel="noopener noreferrer" href="https://duckduckgo.com/?q=!g+caching+background+R+processes&t=ffab&ia=web">
does little to help uncover any useful information
</a>.) Caching is implemented because it saves time, generally by saving the
results of one function call for subsequent reuse. Background processes are also
commonly implemented as time-saving measures, through delegating long-running
tasks to "somewhere else", allowing you to keep focussing on whatever
(un)important things you were doing in the meantime.

Straightforward caching of the results of single function calls is often
achieved through
<a target="_blank" rel="noopener noreferrer" href="https://duckduckgo.com/?q=!w+memoization&t=ffab&ia=web">
"memoisation"
</a>, implemented in several **R** packages including
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/package=R.cache">
R.cache
</a>, 
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/package=memoise">
memoise
</a>, 
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/package=memo">
memo
</a>, 
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/package=simpleCache">
simpleCache
</a>, and
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/package=simpleRCache">
simpleRCache
</a>, not to mention the extremely useful cache-management package,
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/package=hoardr">
hoardr
</a>. None of these packages offer the ability to perform the caching via
a background process, and thus the initial call to a function to-be-cached will
have to wait until that function finishes before returning a value.

This blog entry describes how to implement caching via background processes.
Using a background process to cache an object naturally requires a measure of
anticipation that the object to be cached is likely to be useful sometime in the
future, as opposed to necessarily needed right now. This is nevertheless
a relatively common situation is complex, multi-stage analyses, where the
results of one stage generally proceed in a predictable manner to subsequent
stages. The typical inputs and outputs of those subsequent stages are the things
that can be anticipated, and the results pre-calculated via background
processes, and then cached for subsequent *and immediate* recall. So having
briefly described "standard" caching ("foreground" caching, if you like), it's
time to describe background processes in **R**.

## Background processes in R

Background processes are, among other things, the key to the much-used 
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/package=future">
future package
</a>. This package seems at first like a barely intelligible miracle of
mysterious implementation. What are these "futures"? The host of highly
informative vignettes provide a wealth of information on how the users of this
package can implement their own "futures", yet little information on how the
futures themselves are implemented. (This is not a criticism; it reflects
a reasonably self-justifying design choice, because the average user of this
package will be generally satisfied with knowing how to use the package, and
won't necessarily want or need to know *how* the magic is performed.)

In short: a "future" is just a background process that dumps its results
somewhere ready for later recall. What is a background process? Simply another
**R** session running as a separate 
<a target="_blank" rel="noopener noreferrer" href="https://duckduckgo.com/?q=!w+computer+process&t=ffab&ia=web">
process
</a>. It's easy to implement in base R. We first need a simple **R** script,
as for example generated by the following code:
```{r my_code, eval = TRUE}
my_code <- c ("x <- rnorm (1e6)",
                           "y <- x ^ 2",
                           "y [x < 0] <- -y [x < 0]",
                           "saveRDS (sd (y), file = 'myresult.Rds')")
writeLines (my_code, con = "myfile.R")
```
That script can be executed as a background process by simply calling 
<a target="_blank" rel="noopener noreferrer" href="https://stat.ethz.ch/R-manual/R-devel/library/utils/html/Rscript.html">
Rscript
</a> via a
<a target="_blank" rel="noopener noreferrer" href="https://stat.ethz.ch/R-manual/R-devel/library/base/html/system.html">
system
</a> or
<a target="_blank" rel="noopener noreferrer" href="https://stat.ethz.ch/R-manual/R-devel/library/base/html/system2.html">
system2
</a> call, where the latter two allow `wait = FALSE` to send the process to the
background. (The more recent implementation of system calls via the
<a target="_blank" rel="noopener noreferrer" href="https://github.com/jeroen/sys">
sys package
</a> and its simple `exec_background()` function also deserve a mention here.)
In base R terms, a script can be called from an interactive session via
```{r background, eval = TRUE}
system2 (command = "Rscript", args = "myfile.R", wait = FALSE)
list.files (pattern = "^my")
```
The script has been executed as a background process, and the result dumped to
the file, "myresult.Rds". This can then simply be read to retrieve the cached
result generated by that background process:
```{r}
readRDS ("myresult.Rds")
```

And that value was calculated in, and cached from, a background process. Simple.

### Complications

Where was the above value stored? In the working directory of that **R**
session, of course. This is often neither a practicable nor sensible approach,
for example whenever any control over storage locations is desired. These
cached values are generally going to be temporary in nature, and the
`tempdir()` of the current **R** session offers an alternative location, and is
in fact the only location acceptable for CRAN packages to write to during
package tests. Other common options include a sub-directory of `~/.Rcache`, as used for example in the 
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/package=R.cache">
R.cache
</a> package. I'll only consider `tempdir()` from here on, but doing so will
also reveal why the more enduring location of `~/.Rcache` is often preferred.

Another complication arises in calling 
<a target="_blank" rel="noopener noreferrer" href="https://stat.ethz.ch/R-manual/R-devel/library/utils/html/Rscript.html">
Rscript
</a>, by virtue of the claims in
<a target="_blank" rel="noopener noreferrer" href="https://cran.r-project.org/doc/manuals/r-release/R-exts.html">
"Writing R Extensions"
</a> -- the official CRAN guide to **R** packages -- that one should,

> ... not invoke R by plain R, Rscript or (on Windows) Rterm in your examples,
tests, vignettes, makefiles or other scripts. As pointed out in several places
earlier in this manual, use something like "\$(R_HOME)/bin/Rscript" or "\$(R_HOME)/bin\$(R_ARCH_BIN)/Rterm"

That comment is not very helpful because the alluded "several places" are in
different contexts, and are also only examples rather than actual guidelines. The problem is those suggestions will usually, *but not always* work, depending on Operating System idiosyncrasies. So calling 
<a target="_blank" rel="noopener noreferrer" href="https://stat.ethz.ch/R-manual/R-devel/library/utils/html/Rscript.html">
Rscript
</a> directly is less straightforward than it might seem. 

A further problem arises in that both `system` and `system2` will generally
return values of `0` when everything works okay. "Works" then means that the
process has been successfully started. But where is that process in relation to
the current **R** session? And likely most importantly, has that process
finished or is it still operating? While it is possible to use further `system`
calls to determine the 
<a target="_blank" rel="noopener noreferrer" href="https://duckduckgo.com/?q=!w+process+identifier&t=ffab&ia=web">
process identifier (PID)
</a>, that process itself is fraught and perilous. There are further
complications which arise through directly calling background **R** processes
via `Rscript`, but those should suffice to argue for the fabulous alternative
available thanks to Gábor Csárdi and ...

## The processx package

The
<a target="_blank" rel="noopener noreferrer" href="https://github.com/r-lib/processx">
processx
</a> package states simply that it provides,

> "Tools to run system processes in the background"

This package is designed to run *any* available system process, including ones
that potentially have nothing to do with **R** let alone a current **R**
session. Using 
<a target="_blank" rel="noopener noreferrer" href="https://github.com/r-lib/processx">
processx
</a> to run background **R** process thus requires calling `Rscript`, with the
associated problems described above. Fortunately for us, Gábor foresaw this
need and created the "companion" package, 
<a target="_blank" rel="noopener noreferrer" href="https://github.com/r-lib/callr">
callr
</a> to simply

> "Call R from R"


<a target="_blank" rel="noopener noreferrer" href="https://github.com/r-lib/callr">
callr
</a> relies directly on
<a target="_blank" rel="noopener noreferrer" href="https://github.com/r-lib/processx">
processx
</a>, but provides the far simpler function,
<a target="_blank" rel="noopener noreferrer" href="https://callr.r-lib.org/reference/r_bg.html">
r_bg
</a> to

> "Evaluate an expression in another R session, in the background"

So
<a target="_blank" rel="noopener noreferrer" href="https://callr.r-lib.org/reference/r_bg.html">
r_bg
</a> provides the perfect tool for our needs. This function directly evaluates
R code, without needing to render it to text as we did above in order to write
it to an external script file. An 
<a target="_blank" rel="noopener noreferrer" href="https://callr.r-lib.org/reference/r_bg.html">
r_bg
</a> version of the above would look like this:
```{r my_fn}
f <- function () {
    x <- rnorm (1e6)
    y <- x ^ 2
    y [x < 0] <- -y [x < 0]
    saveRDS (sd (y), file = "myresult.Rds")
}
callr::r_bg (f)
```
We immediately see that 
<a target="_blank" rel="noopener noreferrer" href="https://callr.r-lib.org/reference/r_bg.html">
r_bg
</a> returns a handle to the process itself, along with the single piece of
critical diagnostic information: Whether the process is still running or not:
```{r r_bg}
px <- callr::r_bg (f)
px
Sys.sleep (1)
px
```
Multiple processes can be generated and queried this way. The package is
designed around, and returns, 
<a target="_blank" rel="noopener noreferrer" href="https://github.com/r-lib/R6">
R6
</a> class objects, enabling function calls on the objects, notably including
the following:

```{r r_bg2}
px <- callr::r_bg (f)
px
while (px$is_alive())
    px$wait ()
px
```
The `px$is_alive()` and `px$wait()` functions are all that is needed to wait
until a background process is finished. In the context of using background
processes to cache objects, these lines enable the primary **R** session to
simply wait until caching is finished before retrieving the object.

## processx, callr, and caching

There is only one remaining issue with the above code: Where is "myresult.Rds"
in the following code?
```{r my_fn2}
f <- function () {
    x <- rnorm (1e6)
    y <- x ^ 2
    y [x < 0] <- -y [x < 0]
    saveRDS (sd (y), file = file.path (tempdir (), "myresult.Rds"))
}
px <- callr::r_bg (f)
```
It's in `tempdir()`, but *not* the `tempdir()` of the current process. Where is
his other `tempdir()`? It's temporary of course, so has been dutifully cleaned up,
thereby removing our desired result. What is needed is a way to store the
result in the `tempdir()`of the current -- active -- **R** session. This
`tempdir()` is merely specified as a character string, which we can pass
directly to our function:
```{r myfn3}
f <- function (temp_dir) {
    x <- rnorm (1e6)
    y <- x ^ 2
    y [x < 0] <- -y [x < 0]
    saveRDS (sd (y), file = file.path (temp_dir, "mynewresult.Rds"))
}
```
We then only need to note that the second parameter of
<a target="_blank" rel="noopener noreferrer" href="https://callr.r-lib.org/reference/r_bg.html">
r_bg
</a> is `args`, which is,

> "Arguments to pass to the function. Must be a list."

That is then all we need, so let it run ...
```{r r_bg3}
px <- callr::r_bg (f, list (tempdir ()))
while (px$is_alive())
    px$wait ()
list.files (tempdir (), pattern = "^my")
```
And there is our new result, along with all we need to understand how to cache
objects via background **R** processes.

## Summary

1. Define a function to generate the object to be cached, and include
   a `tempdir()` parameter if that is to be used as the cache location.
2. Use `callr::r_bg()` to call that function in the background and deliver the
   result to the desired location.
3. Examine the handle of the process returned by `r_bg()` to determine whether
   it has finished or not.
4. ... use the cached result.

